---
title: "Discussion 2"
author: "Hanying"
output:
  html_document

---

## Logistics

In our discussion, students are supposed to form random small groups and discuss with each other.


## Check usage of homework auto-grade and submission

Does anyone have questions about them? In particular,

1. Do you know how to check your answer from previous attempt of AG and improve your solution? 

2. A friendly reminder: check the file you submitted to Canvas to ensure that you has included all necessary materials in the file (like the figure).

3. Any issue with homework submission?


## Warm-up

### Plan

Work on the Week 2 Quiz question 6 with your neighbor. Choose the answers you think make sense and explain how you know. 

Three minutes to discuss. Then you can explain your answer and the reason.

### Question

Choose the response that best completes the following statement.

Frequency distributions are used when variables are _______, whereas probability distributions are used when variables are _______.

* continuous; discrete
* discrete; continuous
* interval level; ordinal level
* ratio level, nominal level


### Explanation

Frequency distribution reflects how frequently (how often/how many times) a score occurs, and is used for discrete variables, so that we can count the frequency of each score.

Probability distribution reflects the probability or likelihood that a score will occur, and is used for continuous variables.

## Exercise 1

### 10 minutes for group discussion

Every group can share a single laptop. If you prefer, you can choose to work on your own laptops, too.

### Download discussion materials (Exercise 1(a))

* Create a folder for Discussion 2 (Recommended folder names: Disc2, Discussion2). (Right click > new folder > change the folder name)
* Open Canvas, click `Discussion` under `Week 2: Sep 11 - Sep 17`.
* Click all file links to download them into the file you created. (Click the link > Click "Download" in the upper right corner > choose the folder you created for discussion 2.)

### Open file in RStudio (Exercise 1(b))

* Open RStudio
* In the file panel, open the folder for discussion 2
* Click `02 Disc R.Rmd` to open it
* Change Author to your name
* Set working directory to the current folder: Session > Set Working Directory > To files pane location

### Exercise 1(c) Load data

```{r}
lbw <- read.csv("lbw.csv", header = TRUE)
```

This code means that we used the function `read.csv()` to read the data saved in the file `lbw.csv`. And we created a variable in R named `lbw`, it stores the data.

* Check `lbw` shows in the "environment" panel.
* Click the blue arrow to check the information about it.

### Exercise 1(d)

Try to run `View(lbw)` or click `lbw` in the environment.

## Exercise 2

### 10 minutes for discuss

### 2(a)

* bwt is the birth weight (in grams) for each of the 189 babies observed.
* low is a variable that is 0 if the birth weight is at least 2500 g and 1 if the birth weight is less than
2500 g.
* smoke is a variable that is 0 if mother reported not smoking during pregnancy and 1 if mother reported smoking during pregnancy.

Identify the variable type for each of the three variables based on the description. Check the variable type in R. Does the data type in R agree with the true meaning of these variables?



> The birth weight is numeric. However, `low` and `smooke` are indicators and are categorical variables. They are recognized as integers (int) and thus numeric.

### 2(b)

Run the following code to resave low and smoke as categorical vectors in the lbw data frame.
Notice the $ after the data frame name lbw pulls up a list of all of the columns that are
defined in lbw. And the as.factor() function changes the variable type to categorical.


```{r}
lbw$low <- as.factor(lbw$low)
lbw$smoke <- as.factor(lbw$smoke)
```

### 2(c)

Reference the Environment tab or rerun str(lbw) to confirm low and smoke have been updated correctly.

```{r}
str(lbw)
```


## Exercise 3

Consider comparing the variable for birthweight (bwt) across the groups with smoke = 0 (non smoking) and smoke = 1 (smoking).

### 3a

Create comparative boxplots and histograms of the variable bwt between babies with mothers who had smoking status of 0 = No and 1 = Yes.

### 3a(i)

Save off two data frames, SmokeY and SmokeN, to hold the data for those babies whose mothers smoked and those whose did not

```{r}
SmokeY <- subset(lbw, smoke == "1")
SmokeN <- subset(lbw, smoke == "0")
```


### 3a(ii)

Then save off the two vectors of bwt for those two dataframes into bwtSmokeY and bwtSmokeN.

```{r}
bwtSmokeY <- SmokeY$bwt
bwtSmokeN <- SmokeN$bwt
```

### 3a(iii)

Update the following boxplot code to include labels that show which data is which.


Original code:

```{r}
boxplot(bwtSmokeN, bwtSmokeY,
        main = "Birthweights", xlab = "Birthweight (g)")
```

Add name to the figure:

```{r}
boxplot(bwtSmokeN, bwtSmokeY, names = c("No Smoke", "Smoke"),
        main = "Birthweights", xlab = "Birthweight (g)")
```

### 3a(iv)

Update the following histogram code so that both histgrams have x axis classes from 0 to 5000 with a width of 250.

Original code: 

```{r}
# this makes two rows and 1 column for graphs
par(mfrow = c(1,2))

hist(bwtSmokeY, 
     main = "Smoking Mothers", xlab = "Birthweight")

hist(bwtSmokeN, 
     main = "Non Smoking Mothers", xlab = "Birthweight")

# this makes one row and one column for graphing (restoring to default)
par(mfrow = c(1,1)) 
```

First, let's change the bar width. "A histogram has x axis classes from 0 to 5000 with a width of 250" means the breakpoints for the histogram should be `0, 250, 500, ..., 5000`. First recall how to create such a vector:

```{r}
# start from 0, end at 5000, step length is 250
seq(0, 5000, 250)
```

Use this vector as the breakpoints of the histograms:

```{r}
# this makes two rows and 1 column for graphs
par(mfrow = c(2,1))

hist(bwtSmokeY, breaks = seq(0, 5000, 250),
     main = "Smoking Mothers", xlab = "Birthweight")

hist(bwtSmokeN, breaks = seq(0, 5000, 250),
     main = "Non Smoking Mothers", xlab = "Birthweight")

# this makes one row and one column for graphing (restoring to default)
par(mfrow = c(1,1)) 
```


Also, choose a more useful y axis for both graphs.

Observe the figures, we can see the highest bar is around 15.

```{r}
par(mfrow = c(2,1))

hist(bwtSmokeY, breaks = seq(0, 5000, 250), ylim = c(0, 15),
     main = "Smoking Mothers", xlab = "Birthweight")

hist(bwtSmokeN, breaks = seq(0, 5000, 250), ylim = c(0, 15),
     main = "Non Smoking Mothers", xlab = "Birthweight")

par(mfrow = c(1,1)) 
```

We can see some bars in the second histogram is truncated. So, let's slightly increase the range:

```{r}
# this makes two rows and 1 column for graphs
par(mfrow = c(2,1))

hist(bwtSmokeY, breaks = seq(0, 5000, 250), ylim = c(0, 20),
     main = "Smoking Mothers", xlab = "Birthweight")

hist(bwtSmokeN, breaks = seq(0, 5000, 250), ylim = c(0, 20),
     main = "Non Smoking Mothers", xlab = "Birthweight")

# this makes one row and one column for graphing (restoring to default)
par(mfrow = c(1,1)) 
```

Why is it important that the x and y axis are consistent across the two histograms?



### 3(b)

Compare the center, variability and shape of the two groups’ data using the graphs and numeric summaries. Write your own code to find the numeric summaries. 

Figures we have:


```{r, echo = FALSE}
boxplot(bwtSmokeN, bwtSmokeY, names = c("No Smoke", "Smoke"),
        main = "Birthweights", xlab = "Birthweight (g)")
```


```{r, echo = FALSE}
# this makes two rows and 1 column for graphs
par(mfrow = c(2,1))

hist(bwtSmokeY, breaks = seq(0, 5000, 250), ylim = c(0, 20),
     main = "Smoking Mothers", xlab = "Birthweight")

hist(bwtSmokeN, breaks = seq(0, 5000, 250), ylim = c(0, 20),
     main = "Non Smoking Mothers", xlab = "Birthweight")

# this makes one row and one column for graphing (restoring to default)
par(mfrow = c(1,1)) 
```


#### Compare the center.

Based on the boxplots: non-smoking group has higher center.

Based on the histograms: non-smoking group has a higher mode (around 3000) than that of the smoking group (around 2400).

Check with the numeric summaries:

```{r}
# Calculate mean
mean(bwtSmokeN)
mean(bwtSmokeY)

# Calculate the median
median(bwtSmokeN)
median(bwtSmokeY)
```

The non-smoking group is higher than the smoking group in terms of both mean and median, which agrees with our observation from the figures.

#### Compare the variability

Based on the boxplots: the smoke boxplot is shorter.

Based on the histograms: the smoking group is clustered more in the center.

Check with the numeric summaries:

```{r}
# quantiles
quantile(bwtSmokeN)
quantile(bwtSmokeY)

# standard deciation
sd(bwtSmokeN)
sd(bwtSmokeY)
```
IQR for non-smoking: 3621.5 - 2509.0 = 1112.5.

IQR for smoking: 3245.75 - 2370.50 = 875.25.

(Can directly use `IQR` function)

```{r}
IQR(bwtSmokeN)
IQR(bwtSmokeY)
```


We can see that the smoking group has a smaller standard deviation and IQR, which implies that they has smaller variability, which agrees with our observation from the figures.

#### Shape

Both groups’ data looks roughly symmetric in the histograms and boxplots. We see there is 1 outlying value in the smoking group.

