---
title: "Stat 371 Week 2"
author: "Hanying Jiang"
date: "Sep 2023"
output:
  xaringan::moon_reader:
    css: ["default", "css/xaringan-themer.css"]
    lib_dir: libs
    self_contained: false
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
    seal: false

---

class:title


$\def\Dir{\text{Dir}}$
$\def\E{\mathbb{E}}$
$\def\Mult{\text{Mult}}$
$\def\*#1{\mathbf{#1}}$
$\def\m#1{\boldsymbol{#1}}$
$\def\Unif{\text{Unif}}$
$\def\win{\tilde{w}_{\text{in}}}$
$\def\reals{\mathbb{R}}$
$\def\Gsn{\mathcal{N}}$
$\def\argmin{arg\,min}$
$\newcommand{\wout}{\tilde w_{\text{out}}}$
$\newcommand{\vec}{\textit}$

```{r, echo = FALSE, warnings = FALSE, message = FALSE}
library(RefManageR)
library(knitr)
library(tidyverse)
library(glue)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE, dpi = 200, fig.align = "center", fig.width = 6, fig.height = 3, eval = TRUE)
opts_knit$set(eval.after = "fig.cap")
set.seed(20230120)

# BibOptions(cite.style = "authoryear")
# bib <- ReadBib("ref.bib")
```


.center[
<br/>
## Stat 371 Week 2
<br/>
<br/>
]


.pull-right[
Sep 2023

Hanying Jiang


[hjiang252@wisc.edu](hjiang252@wisc.edu)
]

---

### Logistics

In our discussion, students are supposed to form random small groups and discuss with each other.



---

### Check usage of homework auto-grade and submission

Does anyone have questions about them? In particular,

1. Do you know how to check your answer from previous attempt of AG and improve your solution? 

2. A friendly reminder: check the file you submitted to Canvas to ensure that you has included all necessary materials in the file (like the figure).

3. Any issue with homework submission?

---

### Warm-up


Work on the Week 2 Quiz question 6 with your neighbor. Choose the answers you think make sense and explain how you know. 

Three minutes to discuss. Then you can explain your answer and the reason.

---

### Exercise 6

Choose the response that best completes the following statement.

Frequency distributions are used when variables are _______, whereas probability distributions are used when variables are _______.

* continuous; discrete
* discrete; continuous
* interval level; ordinal level
* ratio level, nominal level

---

### Exaplanation

Frequency distribution reflects how frequently (how often/how many times) a score occurs, and is used for discrete variables, so that we can count the frequency of each score.

Probability distribution reflects the probability or likelihood that a score will occur, and is used for continuous variables.

---

### Discussion: try to solve question 1 with your group

Every group can share a single laptop. If you prefer, you can choose to work on your own laptops, too.

---

### Download discussion materials (Exercise 1(a))

* Create a folder for Discussion 2 (Recommended folder names: Disc2, Discussion2). (Right click > new folder > change the folder name)
* Open Canvas, click `Discussion` under `Week 2: Sep 11 - Sep 17`.
* Click all file links to download them into the file you created. (Click the link > Click "Download" in the upper right corner > choose the folder you created for discussion 2.)


---

### Open file in RStudio (Exercise 1(b))

* Open RStudio
* In the file panel, open the folder for discussion 2
* Click `02 Disc R.Rmd` to open it
* Change Author to your name
* Set working directory to the current folder: Session > Set Working Directory > To files pane location

---

### Exercise 1(c) Load data

```{r}
lbw <- read.csv("lbw.csv", header = TRUE)
```

This code means that we used the function `read.csv()` to read the data saved in the file `lbw.csv`. And we created a variable in R named `lbw`, it stores the data.

* Check `lbw` shows in the "environment" panel.
* Click the blue arrow to check the information about it.

---

### Exercise 1(d)

Try to run `View(lbw)` or click `lbw` in the environment.

---

### Discuss Exercise 2

---


### 2(a)

Identify the variable type for each of the three variables based on the description.

> The birth weight is numeric. However, `low` and `smooke` are indicators and are categorical variables. They are recognized as integers (int) and thus numeric.

---

### 2(b)(c)

Run the following code. and pay attention to the usage of `$` sign.


```{r}
lbw$low <- as.factor(lbw$low)
lbw$smoke <- as.factor(lbw$smoke)
str(lbw)
```

---

### Discuss about Exercise 3

---

### 3(a)

```{r}
SmokeY <- subset(lbw, smoke == "1")
SmokeN <- subset(lbw, smoke == "0")
```


### 3(b)

```{r}
bwtSmokeY <- SmokeY$bwt
bwtSmokeN <- SmokeN$bwt
```

### 3(c)

```{r, eval = FALSE}
boxplot(bwtSmokeN, bwtSmokeY, names = c("No Smoke", "Smoke"),
        main = "Birthweights", xlab = "Birthweight (g)")
```


```{r}

```










